Step by Step: Using Step Selection Analysis to Simulate Dispersal and Assess Landscape Connectivity

### 2021-04-22 Chapter_1/04_Manuscript/Manuscript.tex
Worked on manuscript for 1 chap

\documentclass[abstract=on,10pt,a4paper,bibliography=totocnumbered]{article}
\usepackage[paper=a4paper,left=35mm,right=35mm,top=25mm,bottom=30mm]{geometry}
\usepackage[doublespacing]{setspace}
\usepackage[english]{babel}
\usepackage[utf8]{inputenc}
\usepackage[round]{natbib}
\usepackage{amsmath}
\usepackage{colortbl}
\usepackage{amsfonts}
\usepackage{amssymb}
\usepackage{gensymb}
\usepackage{graphicx}
\usepackage{tikz}
\usepackage{enumerate}
\usepackage{enumitem}
\usepackage{subcaption}
\usepackage{booktabs}
\usepackage[hidelinks]{hyperref}
\usepackage[nameinlink]{cleveref}
% \usepackage{lineno}
\usepackage{multirow}
\usepackage{arydshln}
% \usepackage[nomarkers, nolists]{endfloat}

@@

In a previous paper, we assessed landscape connectivity within the KAZA-TFCA for
dispersing African wild dogs using a least-cost corridor approach. For this, we
fitted a basic habitat selection model based on which we predicted landscape
permeability. We now expand on this knowledge and develop a more detailed
movement model of dispersing wild dogs. We then use this model to simulate
thousands of dispersers moving throughout the KAZA. Based on said simulations,
we compute heatmaps and identify potential dispersal hotspots and compare them
to the dispersal routes identified in \citep{Hofmann.2021}. We also showcase how
simulated dispersal data can be analysed using network-analysis and how network
metrics relevant to landscape connectivity can be computed. Our results suggest
that a simulation based approach yields several major benefits over traditional
connectivity modelling techniques. Most importantly, simulations provide a more
generic view on how connectivity emerges and to which degree connectivity
depends on the dispersal duration. In addtion, by generating proper dispersal
trajectories, network theory can be applied to calculate network metrics that
are pertinent to connectivity analysis. Finally, we put forward additional
opportunities using simulations that go beyond the scope of this paper.

@@

\section{Methods}
\subsection{Study Area}
The study area (centered at -17\degree 13'9''S, 23\degree 56'4''E;
\Cref{StudyArea}a) stretched over 1.3 Mio km\textsuperscript{2} and ecompassed
the entire KAZA-TFCA (\Cref{StudyArea}b). The KAZA-TFCA is the world's largest
transboundary conservation area and comprises parts of Angola, Botswana,
Namibia, Zimbabwe, and Zambia, covering a total area of over 520'000
km\textsuperscript{2}. Its landscape varies regionally and ranges from savanna,
to grassland, and from dry to moist woodland habitats. A dominant
hydrogeographical feature in its center is the Okavango Delta, the earth's
largest inland delta. The delta and its surroundings are considered a stronghold
for African wild dogs and may act as a source for the recolonization of
surrounding habitats. The wet season within our study area lasts from November
to March and is out of phase with the main flooding of the Okavango Delta which
peaks between July and August \citep{McNutt.1996, Wolski.2017}.

\begin{figure}[htbp]
  \begin{center}
    \includegraphics[width = \textwidth]{99_StudyArea.png}
    \caption{Study area}
    \label{StudyArea}
  \end{center}
\end{figure}

\subsection{GPS Relocation Data}
Between 2011 and 2020, we collected GPS relocation data on dispersing wild dogs
from a free-ranging wild dog population inhabiting the Okavango Delta in
northern Botswana. We identified potential dispersers based on age, number of
same‐sex siblings, pack size, and presence of unrelated opposite-sex individuals
in the pack \citep{McNutt.1996, Behr.2020}. We immobilized individuals using a
cocktail of Ketamine/Xylazine/Atropine \citep{Osofsky.1996, Cozzi.2020},
injected with a dart that was fired from a CO\textsubscript{2}-pressurized gun
(\textit{DAN-Inject, Denmark}). Protocols for the anesthization protocols were
approved by the Ministry of Environment, Natural Resources Conservation and
Tourism of Botswana (permit EWT 8/36/4 XXXVI). After immobilization, individuals
were fitted with GPS/Satellite radio collars (\textit{Vertex Lite; Vectronic
Aerospace GmbH, Berlin}) that included an automated drop-off mechanism. Handling
and collaring of all individuals was carried out and supervised by a
Botswana-registered wildlife veterinarian. All of the immobilized individuals
quickly rejoined their pack within one hour after the procedure. Out of all
collared individuals, 16 eventually dispersed in separate same-sex coalitions
and their trajectories were successfully recorded (7 female and 9 male
coalitions).

During dispersal, collars were programmed to record a GPS fix every 4 hours.
Recorded relocations were regularly transmitted over the Iridium satellite
system, which allowed remote tracking of individuals, even if they left the main
study area and crossed international borders. To effectively distinguish between
periods of residency and dispersal, we applied the net-squared displacement
metric to the observed movement data. This metric measures the squared Euclidean
distance of a collared individual to a reference point \citep{Borger.2012},
which we set at the center of each disperser's natal home range. Hence,
dispersal was deemed to have started when an individual left its natal home
range and ended when the individual became stationary again. For the purpose of
this study, we discarded andy data that was collected during residency. Because
previous research could not find any differences between males and females
during dispersal \citep{Woodroffe.2019, Cozzi.2020}, we did not distinguish
sexes for the sake of our analyses. Ultimately, we converted the collected GPS
relocations to steps, where each step represented the straight-line distance
travelled between two consequtive GPS relocations \citep{Turchin.1998}.

\subsection{Covariates}
To represent habitat covariates, we prepared a set of spatial raster layers
depicting water-cover (dynamically updated), distance to water (dynamically
updated), tree-cover, and shrub/grassland-cover. To render a decreasing impact
with increasing distance, we square rooted the distance to water covariate
layer. We also created a proxy for human influence, rendering anthropogenic
pressures stemming from human-density, agricultural sites, and roads. We
rendered all layers at a resolution of 250m by 250m for the entire extent of the
KAZA-TFCA. Further details on the derivation and preparation of each
environmental covariate is given in \cite{Hofmann.2020}.

Besides habitat covariates, we also computed movement metrics that we used as
movement covariates in our models. These covariates included the step length
(\(sl\)), its natural logarithm (\(log(sl)\)), as well as the cosine of the
relative turning angle (\(cos(ta)\)). Because wild dogs follow a diurnal
activity pattern, we furthermore coded a binary variable indicating whether an
observed step was realized during main activity (xx to xx) or low activity (xx
to xx).

\subsection{Movement Model}
We used integrated step selection functions (iSSF) to parametrize a mechanistic
movement model of dispersing wild dogs. In this framework, observed steps are
contrasted to random steps that the animal could have realized, but decided not
to. To compare steps, it is assumed that animals assign a relative selection
score \(w(x)\) to each step. The score depends on the covariates \(X\)
experienced along a step, as well as the animals preferences \(\beta\) towards
these covariates. The probability of a step being realized is then contingent on
the step's selection score, as well as the selection scores of all other step in
the same stratum. In contrast to regular step selection analysis,
\textit{integrated} step selection analysis allows simulatneous inference on
movement and habitat preferences of the studied animal. Moreover, potential
interactions between habitat and movement preferences can be included to
investigate how movement behavior depends on habitat characteristics. In result,
the method produces less biased selection estimates and allows to apply the
resulting model as a fully mechanistic movement model based on which movement
can be simulated. To conduct iSSF analysis, we paired each observed step with 24
random steps. Random steps basically resembled potential alternatives that the
animal could have realized but decided not to. We generated random steps by
sampling random turning angles from a uniform distribution (\(-\pi, +\pi\)) and
step lengths from a gamma distirbution that was fitted to observed steps (scale
= 6'308, shape = 0.37). While the number of random steps is inversely
proportional to the sampling error \citep{Avgar.2016}, we found only minor
differences in model performance when sampling additional random steps and
deemed 24 random steps sufficient. Along each step, we extracted spatial
covariates using the \textit{velox} package and calculated the movement metrics
\(sl\), \(log(sl\)), and \(cos(ta)\). To facilitate model convergence, we
standardized all continuous covariates to a mean of zero and a standard
deviation of one. We then used the r-package \textit{glmmTMB} to fit mixed
effects conditional logistic regression models as proposed by \citep{Muff.2020}.

Our movement model was and extension of the habitat model presented in Hofmann
et al 2021. We will refer to it as the base model. Because the base model was
used to predict landscape permeability, no interactions between the habitat and
movement metrics were modeled. Thus, we expanded the base model and by proposing
additional interactions between movement metrics and environmental covariates.
For this, we started with the base model and iteratively increased model
complexity by adding all proposed two-way interactions between habitat
covariates and movement covariates. For instance, for the covariate
\textit{water} we proposed the interactions \(Water:log(sl)\),
\(Water:log(sl)\), and \(Water:cos(ta)\). Furthermore, we proposed the
interactions \(sl:cos(ta)\) and \(log(sl):cos(ta)\) to account for a correlation
between turning angles and step lengths, as well as the interactions
\(sl:MainActivity\) and \(log(sl):MainActivity\) to account for the fact that
step lengths diffuer due to the diurnal activity pattern. We then ran stepwise
model forward selection based on Akaike's Information Criterion (AIC,
\citealp{Burnham.2002}) values and identified the most parsimonious movement
model. Although several models received an AIC weight above one, we decided to
only consider the  ``best'' model for simplicity. In either way, the models with
positive weights contained almost identical covariates and an averaged model
would have only given weak support to additional covariates.

\subsection{Source Points}
We randomly placed source points within protected areas larger (\(>\) 700
km\textsuperscript{2}), which conforms to the average home range requirement of
resident wild dogs \citep{Pomilia.2015}. By distributing source points randomly,
the number of source points per km\textsuperscript{2} within protected areas was
approximately equal across our study area. In total, we distributed 50'000
source points across all protected areas, each representing the starting point
of a dispersal trajectory. To render potential immigrants into the study system,
we randomly placed 10'000 additional source points inside a xx km buffer around
the study area. Hence, a total of 60'000 source points were generated, thereby
resulting in 60'000 simulated dispersers. It is worth pointing out that our
choice of  placing source points completely random is arbitrary, as one as well
adjust the sampling frequency based on the density of dispersers in the
respective source area.

% Simulations from step selection functions are known to be sensitive to the
% location of source points, which is why we followed a twofold approach to sample
% release points across the KAZA-TFCA. In a first approach, we used the same 68
% source points between which we already computed least-cost paths and least-cost
% corridors (see xx). These source points were regularly spaced 100 km apart,
% located within protected areas (\(>\) 700 km\textsuperscript{2}). Because of
% the regulary spacing, not all protected areas received source points. Hence, we
% placed additional source points at the center of such protected areas. At each
% of the generated source point we released 1000 dispersers, implying a total of
% 68'000 simulated dispersers. In a second approach, we used the same 68 source
% points to define catchment areas inside protected areas. That is, for each of
% the 68 source points we defined its voronoi polygon within the point's protected
% area. Within these catchment areas we then randomly sampled 1000 new source
% points from which we released another 68'000 dispersers. Overall, we simulated
% 134'000 dispersers for a total of 268 Mio. steps.

\begin{figure}[htbp]
  \begin{center}
    \includegraphics[width = \textwidth]{99_SourcePoints.png}
    \caption{Illustration of source points from which dispersal was simulated.
    Although in reality we simulated 1'000 per source point, we illustrate an
    example assuming 10 dispersers from each source point. (a) Static source
    points similar to the source points reported in ... (b) Source points that
    were randomized within the catchment areas (dark gray, delinated by solid
    black lines).}
    \label{SourcePoints}
  \end{center}
\end{figure}

\subsection{Dispersal Simulation}
We used the most parsimonious movement model to simulate a total 80'000
dispersing wild dogs across the KAZA-TFCA. The simulation basically resembled an
inverted iSSF and was set up as follows. First, we sampled a source point at
which a disperser was initiated and assumed a random initial orientation of the
animal. The orientation was essential to permit calculation of relative turning
angles. Second, we generated a set of 25 random steps by sampling turning angles
from a uniform distribution (\(-\pi, +\pi\)) and step lengths from the fitted
gamma distribution. Similarly to our input data, each step corresponded to a
temporal difference of 4 hours. To prevent unrealistically large steps, we
capped sampled step lengths at 35km, which corresponds to the farthest distance
travelled by one of our dispersers within 4 hours. Third, along each random step
we extracted habitat covariates and calculated movement covariates. Fourth, we
applied the parametrized movement model to predict the selection scores \(w(x)\)
for all steps. Fifth, we translated selection scores into probabilities by
applying \Cref{EQ2}. Sixth, we sampled one of the random steps based on assigned
probabilities and determined the animal's new position. We then repeated steps
two to six until a total of 2'000 steps were realized.

Inevitably, some simulated dispersers would approach a map boundary and one of
the generated random steps would leave the study area. However, beyond the study
area no covariates could be extracted and no selection scores can be calculated.
To mitigate the influence of such edge effects, we buffered all covariate layers
by xx percent and filled the buffer zone with covariate values resampled from
the observed data. As such, simulated dispersers were allowed to leave and
reenter the study area through this buffer zone. In case one of the proposed
random steps left the buffer, we resampled the corresponding step until it lied
fully within the buffer zone. This implies that we forced the disperser to
bounce off the borders of the buffer zone.

To meaningfully select source points, we randomly sampled coordinates within
protected areas larger (\(>\) 700 km\textsuperscript{2}), which conforms to the
average home range requirement of resident wild dogs \citep{Pomilia.2015}. By
distributing source points randomly, the number of source points per
km\textsuperscript{2} within protected areas was approximately equal. In total,
we distributed 50'000 source points across protected areas. To render potential
immigrants into the study area, we placed 10'000 additional random points inside
a xx km buffer around the study area. In total, we generated 80'000 source
points, thereby resulting in 80'000 simulated dispersers, each of which moved
2'000 steps. Overall, this implied 160 Mio. simulated steps.

\subsection{Heatmaps}
Using the simulated trajectories, we created heatmaps that indicate the
frequency at which each raster-cell in the study area was visited. If the same
trajectory crossed a pixel twice, it was only counted once, i.e. we did not
consider revisits. We achieved high performance rasterization of all simulated
trajectories using the recently developed R-package \textit{terra}
\citep{Hijmans.2020}.

To examine if and how ``heat'' changes in response to changes in the location of
source points and the number of simulated steps, we followed a 2 x 6 design and
created heatmaps for both point sampling regines, as well as for 68, 125, 250,
500, 1000, and 2000 dispersal steps. We quantified the similarity of the
resulting 12 heatmaps to the permeability and least-cost corridor maps presented
in (Hofmann ...) we used Bhattacharyya's affinity. Bhattacharyya's affinity
ranges from zero (complete separation) to one (perfect match) and has earlier
been proposed to compare the overlap of utilisation distributions (Fieberg).

\subsection{Network Analysis}
We coerced all simulated trajectories into a network consisting of vertices
(relocations) and edges (connections between relocations). To do so, we created
raster layers at multiple resolutions and identified each trajectory's
transition matrix on these rasters see figure xx \citep{BastilleRousseau.2018}.
We then merged the transition matrices of all trajectories and calculated
cumulative transitions between all raster-cells. This resulted in an edge-list,
containing all observed from-to connections as well as their frequency. Using
this edge-list, we generated a weighted graph using the r-package
\textit{igraph}. Based on this graph we calculated betweenness scores as well as
the degree of each raster cell. Betweenness indicates how often a specific
raster-cell lies on a shortest path between two other raster-cells and is a
useful metric to detect movement corridors. Degree, on the other hand, indicates
how many connections a raster cell has and therefore serves to illustrate the
xxx of specific nodes. When calculating betweenness, we used the transition
frequency as weighting factor. That is, a higher transition frequency
contributed to a higher betweenness score.

% Based on the simulated trajectories we identified to which other areas each
% source area is connected. For instance, if a trajectory originated at source
% point one and intersected with source areas two and three, we assumed that
% source areas two and three were within reach of source are one. To get a sense
% of the strength of connections, we also calculated how often each of these
% connections was realized. This procedure resulted in 6 x 2 edge lists which we
% further used to generate weighted networks and to calculate network metrics.

\section{Results}
\subsection{Movement Model}
Compared to the base model, the most parsimonious movement model included
several additional interactions (\Cref{MovementModel} and Table S1). The model
indicates that dispersers move directional, particularly when distant to water,
yet less so in human dominated landscapes. Furthermore, dispersers prefer large
steps, especially during main activity. In contrast, step lengths tend to be
shorter when water- or tree-cover is high. In general, dispersers avoid water,
prefer proximity to water, avoid dense tree-cover, prefer shrubs/grassland, and,
finally, avoid human dominated landscapes.

\begin{figure}
  \begin{center}
    \includegraphics[width=\textwidth]{99_MovementModel}
    \caption{(a) Most parsimonious movement model for dispersing wild dogs. The
    model includes a habitat kernel, a movement kernel, as well as their
    interactions. The orange and blue line segments delineate the 90\%, 95\%,
    and 99\% Confidence-Intervals around the respective \(\beta\) coefficients.
    Significance codes: * \(p < 0.10\), ** \(p < 0.05\), *** \(p < 0.01\). (b)
    Results from the k-fold cross validation procedure. The upper plot shows
    rank frequencies of predicted realized scroes according to model predictions
    with known preferences, whereas the lower plot shows rank frequencies when
    assuming random preferences. The blue ribbon shows the prediction interval
    around a loess smoothing regression that we fitted to ease the
    interpretation of the plots.}
    \label{MovementModel}
  \end{center}
\end{figure}

\subsection{Dispersal Simulation}

\subsection{Heatmaps}
Six of the twelve rasterized dispersal trajectories are presented as heatmaps in
\Cref{Heatmaps}. As can be seen, differences that stem from the method of point
sampling disappear as more steps are simulated. This is to be expected as the
influence of the origin becomes smaller and smaller as the animal moves through
the landscape. However, there are striking differences when simualtions are only
run for few iterations.

\begin{figure}
  \includegraphics[width=\textwidth]{99_Heatmaps}
  \caption{}
  \label{Heatmaps}
\end{figure}

>>>>

\documentclass[abstract=on,10pt,a4paper,bibliography=totocnumbered]{article}
\usepackage[paper=a4paper,left=35mm,right=35mm,top=25mm,bottom=30mm]{geometry}
\usepackage[doublespacing]{setspace}
\usepackage[english]{babel}
\usepackage[utf8]{inputenc}
\usepackage[round]{natbib}
\usepackage{amsmath}
\usepackage{colortbl}
\usepackage{amsfonts}
\usepackage{amssymb}
\usepackage{gensymb}
\usepackage{graphicx}
\usepackage{tikz}
\usepackage{enumerate}
\usepackage{enumitem}
\usepackage{subcaption}
\usepackage{booktabs}
\usepackage[hidelinks]{hyperref}
\usepackage[nameinlink]{cleveref}
% \usepackage{lineno}
\usepackage{multirow}
\usepackage{arydshln}
\usepackage[flushleft]{threeparttable}
\usepackage[nomarkers, nolists]{endfloat}

@@

In a previous paper, we assessed landscape connectivity within the KAZA-TFCA for
dispersing African wild dogs using a least-cost corridor approach. For this, we
fitted a basic habitat selection model based on which we predicted landscape
permeability. We now expand on this knowledge and develop a more detailed
movement model of dispersing wild dogs. We then use this model to simulate
thousands of dispersers moving across the KAZA-TFCA. Based on said simulations,
we compute heatmaps and identify potential dispersal hotspots and compare them
to the dispersal routes identified in \citep{Hofmann.2021}. We also showcase how
simulated dispersal data can be analysed using network-analysis and how network
metrics relevant to landscape connectivity can be computed. Our results show
that a simulation based approach yields several major benefits over traditional
connectivity modelling techniques. Most importantly, simulations provide a more
generic view on how connectivity emerges and to which degree connectivity
depends on the dispersal duration. In addtion, by generating proper dispersal
trajectories, network theory can be applied to calculate network metrics that
are pertinent to connectivity analysis. Finally, we put forward additional
opportunities using simulations that go beyond the scope of this paper.

@@

\section{Methods}
\subsection{Study Area}
The study area was centered at -17\degree 13'9''S, 23\degree 56'4''E
(\Cref{StudyArea}a) and was represented by a rectangular bounding box that
stretched over 1.3 Mio. km\textsuperscript{2}. Its extent encompassed the entire
KAZA-TFCA (\Cref{StudyArea}b), which is the world's largest transboundary
conservation area and comprises parts of Angola, Botswana, Namibia, Zimbabwe,
and Zambia, covering a total of 520'000 km\textsuperscript{2}. Its landscape
varies regionally and ranges from savanna, to grassland, and from dry to moist
woodland habitats. A dominant hydrogeographical feature in its center is the
flood-pulsing Okavango Delta, the earth's largest inland delta. The delta and
its surroundings are considered a stronghold for African wild dogs and may act
as a source for the recolonization of surrounding habitats \citep{Cozzi.2013}.
The wet season within the study area lasts from November to March and is out of
phase with the main flooding of the Okavango Delta which peaks between July and
August \citep{McNutt.1996, Wolski.2017}. While large portions of land within the
KAZA-TFCA are designated protected areas or national parks that remain largely
free from humans, substantial human influence originates from roads,
agricultural sites and several small to large settlements.

\begin{figure}[htbp]
  \begin{center}
    \includegraphics[width = \textwidth]{99_StudyArea.png}
    \caption{(a) The study area of our research was confined by a bounding box
    encompassing the entire KAZA-TFCA, which comprises parts of Angola, Namibia,
    Botswana, Zimbabwe, and Zambia. (b) The KAZA-TFCA represents the world's
    largest terrestrial conservation area and covers a total of 520'000
    km\textsuperscript{2}. It will connect or reconnect multiple already
    existing national parks and protected areas (green polygons) that are
    distributed across its extent. The African wild dog dispersal data
    considered in this study was collected on free-ranging individuals departing
    from the Moremi National Park in northern Botswana (since game reserves in
    Botswana serve the same purpose as national parks, we refer to them as
    national parks for simplicity).}
    \label{StudyArea}
  \end{center}
\end{figure}

\subsection{GPS Relocation Data}
Between 2011 and 2019, we collected GPS relocation data from dispersing wild
dogs in a free-ranging wild dog population inhabiting the Moremi National Park
in northern Botswana. We identified potential dispersers based on age, number of
same‐sex siblings, pack size, and presence of unrelated opposite-sex individuals
in the pack \citep{McNutt.1996, Behr.2020}. We immobilized individuals using a
cocktail of Ketamine/Xylazine/Atropine \citep{Osofsky.1996, Cozzi.2020},
injected with a dart, fired from a CO\textsubscript{2}-pressurized gun
(\textit{DAN-Inject, Denmark}). Anesthesia protocols were approved by the
Ministry of Environment, Natural Resources Conservation and Tourism of Botswana
(permit EWT 8/36/4 XXXVI). After immobilization, individuals were fitted with
GPS/Satellite radio collars (\textit{Vertex Lite; Vectronic Aerospace GmbH,
Berlin}) that included an automated drop-off mechanism. Handling and collaring
of all individuals was carried out and supervised by a Botswana-registered
wildlife veterinarian. All of the immobilized individuals usually rejoined their
pack within one hour after the procedure. Out of all collared individuals, 16
individuals eventually dispersed in separate same-sex coalitions and their
trajectories were successfully recorded (7 female and 9 male coalitions).

During dispersal, collars were programmed to record a GPS fix every 4 hours.
Collected relocations were regularly transmitted over the Iridium satellite
system, which allowed remote tracking of individuals, even if they left the main
study area and crossed international borders. To distinguish between periods of
residency and dispersal, we applied the net-squared displacement metric to the
observed movement data. This metric measures the squared Euclidean distance of a
collared individual to a reference point \citep{Borger.2012}. As a reference
point, we used the center of each disperser's natal home range, such that
dispersal was deemed to have started when an individual left its natal home
range and ended once individuals became sedentary again. For the purpose of this
study, we discarded all data that was collected during residency and only
retained movement data during dispersal. Previous research suggests that females
and males behave similarly during dispersal \citep{Woodroffe.2019, Cozzi.2020},
so we did not distinguish between sexes in our analyses. Ultimately, we
converted the collected GPS coordinates to steps, where each step represented
the straight-line distance traveled by and individual between two consecutive
GPS relocations \citep{Turchin.1998}.

\subsection{Covariates}
We represented the physical landscape in the study area using a set of habitat
covariates rendering water-cover, distance to water, tree-cover, and
shrub/grassland-cover. Because water cover greatly changes within and between
years around the Okavango Delta, we applied a remote sensing technique that
allowed us to generate frequently updated water cover layers and corresponding
distance to water layers (see \citealp{Wolski.2017} and \citealp{Hofmann.2021}).
Hence, the obtained layers temporally aligned with our dispersal events.
Furthermore, we computed a proxy for human influence, depicting anthropogenic
pressures stemming from human-density, agricultural sites, and roads. We
coarsened or interpolated all layers to a target resolution of 250 m by 250 m
and cropped them to the extent of the KAZA-TFCA. Further details on the sources
and preparation of each habitat covariate are given in \cite{Hofmann.2020}.

Besides habitat covariates, we also computed movement metrics that we used as
movement covariates in our models. Movement metrics were calculated for each
step and included the step length (\textsf{sl}), its natural logarithm
(\textsf{log(sl)}), and the cosine of the relative turning angle
(\textsf{cos(ta)}). Because wild dogs follow a diurnal activity pattern, we also
coded a binary variable (\textsf{LowActivity}) indicating whether a step was
realized during low activity (17:00 to 07:00 local time) or high activity (09:00
to 17:00 local time). We conducted all cleaning and analysis, as well as all
simulations using the statistical software {\tt R}, version 3.6.6
\citep{R.2019}. The corresponding {\tt R}-scripts are provided on Github.

\subsection{Movement Model}
We used integrated step selection functions (iSSF) to parametrize a mechanistic
movement model of dispersing wild dogs. In contrast to regular step selection
functions, \textit{integrated} step selection functions allow simultaneous
inference on movement and habitat preferences of the studied animal, and on
potential interactions between habitat and movement preferences. In result, the
method produces less biased selection estimates and allows to treat the
resulting model as a fully mechanistic movement model based on which movement
can be simulated. To conduct iSSF analysis, we paired each observed step with 24
random steps, alltogether forming a stratum that received a unique identifier.
Each of the 24 random steps served as pseudeo-absence and resembled a potential
alternative that the animal could have realized but decided not to. We generated
random steps by sampling random turning angles from a uniform distribution
(\(-\pi, +\pi\)) and step lengths from a gamma distirbution that was fitted to
observed steps (scale = 6'308, shape = 0.37). While the number of random steps
is inversely proportional to the sampling error \citep{Avgar.2016}, a
preliminary analysis revealed only minor differences in model performance when
sampling additional random steps, hence we deemed 24 random steps to be
sufficient. Along each step, we extracted and averaged spatial covariates using
the {\tt velox} package and we calculated the movement metrics \textsf{sl},
\textsf{log(sl)}, and \textsf{cos(ta)}. To facilitate model convergence, we
standardized all continuous covariates to a mean of zero and a standard
deviation of one. Since correlation among covariates was low (\(|r| > 0.6\);
\citealp{Latham.2011}), we retained all of them for modeling.

To compare realized steps (scored 1) to random steps (scored 2), we assumed that
animals assigned a selection score \(w(x)\) to each step. The score depended on
the step's covariates \(X\), as well as the animals preferences \(\beta\)
towards these covariates. The probability of a step being realized was then
contingent on the step's selection score, as well as on the selection scores of
all other step in the same stratum. To estimate the preferences of interest, we
fitted conditional logistic regression models in the r-package {\tt glmmTMB}
following the method proposed by \citep{Muff.2020}. This approach enables to fit
random slopes in addition to random intercepts. Here, we used animal IDs as
random effect and modelled random slopes for each covariate. To apply the
``poisson trick'', we fixed the random intercept variance at an arbitrary high
value of 10\textsuperscript{6} \citep{Muff.2020}.

The structure of our movement model was based on a habitat selection model for
dispersing wild dogs presented in \cite{Hofmann.2021}. In the original model, no
interactions between the habitat and movement covariates were considered, so we
slightly expanded this base model by proposing interactions between movement and
habitat covariates. More specifically, we started with the base model and
incrementally increased model complexity by adding all possible two-way
interactions between habitat covariates and movement covariates. For instance,
for the covariate \textsf{water}, we proposed the interactions
\textsf{Water:log(sl)}, \textsf{Water:log(sl)}, and \textsf{Water:cos(ta)}.
Besides those combinations, we also proposed the interactions
\textsf{sl:cos(ta)} and \textsf{log(sl):cos(ta)} to account for a correlation
between turning angles and step lengths, as well as the interactions
\textsf{sl:LowActivity} and \textsf{log(sl):LowActivity} to account for the fact
that step lengths may differ due to wild dogs' diurnal activity pattern. To
compare competing models and assess the most parsimonious movement model, we ran
stepwise forward model selection based on Akaike's Information Criterion (AIC,
\citealp{Burnham.2002}) and computed model weights.

We validated the predictive power of the selected movement model using a k-fold
cross-validation procedure for case-control studies presented in
\cite{Fortin.2009}. We randomly split the input data into training and testing
data with an 80:20 ratio. Using the training data we parametrized a movement
model based on which we predicted selection scores \(w(x)\) for all steps in the
test data. Within each stratum we then assigned ranks 1-25 to each step based on
predicted selection scores such that rank 1 was given to the step with the
highest score \(w(x)\). Across all strata we determined the realized step's rank
and calculated rank frequencies for realized steps. Finally, we computed
Spearman's rank correlation between ranks and associated frequencies \(r_{s,
realized}\). We replicated the entire procedure 100 times and computed the mean
correlation coefficient (\(\bar{r}_{s, realized}\)), as well as its 95\%
confidence interval across all replicates. For comparison, we repeated the same
procedure 100 times assuming random preferences. Random preferences were
implemented by discarding the realized step from all strata and identifying the
rank of a random step in each stratum. Again, we calculated Spearman's rank
correlation coefficient (\(r_{s, random}\)), its mean across repetitions
(\(\bar{r}_{s, random}\)), and its 95\% confidence interval. This validation
proofs a significant prediction in case the confidence intervals of
\(\bar{r}_{s, realized}\) and \(\bar{r}_{s, random}\) do not overlap.

\subsection{Dispersal Simulation}
We used the most parsimonious movement model to simulate a total of 80'000
dispersing wild dogs moving across the KAZA-TFCA. The simulation resembled an
inverted iSSF and was set up as follows. (1) We sampled a source point at which
a disperser was initiated. We assumed a random initial orientation of the animal
in order to permit calculation of relative turning angles. (2) We generated a
set of 25 random steps by sampling turning angles from a uniform distribution
(\(-\pi, +\pi\)) and step lengths from the fitted gamma distribution. In line
with our training data, each step resembled the straight line movement realized
within 4 hours. To prevent the sampling of unreasonably large steps, we capped
sampled step lengths to a maximum of 35 km, which corresponds to the farthest
distance ever traveled within 4 hours by our dispersers. (3) Along each random
step, we extracted habitat covariates and calculated movement covariates. (4) We
applied the parameterized movement model to predict the selection score \(w(x)\)
for each of the proposed steps and translated selection scores into
probabilities by applying \Cref{EQ2}. (5) We sampled one of the random steps
based on predicted probabilities and used it to determine the animal's new
position. We then repeated steps (2) to (5) until 2'000 steps were realized.

Because some simulated dispersers would inevitably approach a map boundary, some
of the proposed random steps may leave the study area, such that covariates
could be extracted and no selection score could be calculated. To mitigate the
influence of such edge effects, we followed \citep{Koen.2010} and artificially
expanded all covariate layers by adding a 100 km buffer zone within which we
randomized covariate values by resampling values from the observed covariate
layers. This allowed simulated dispersers to leave and reenter the study area
through this buffer zone. Nevertheless, if a disperser approached this outer
boundary and one of the proposed random steps left the buffer, we resampled
transgressing steps until they fully lied within the buffer zone, thereby
forcing dispersers to bounce off the outer boundary. The simulation was mainly
coded in the programming language {\tt R}, yet a few helper functions were
written in {\tt C++} and imported into {\tt R} using the {\tt Rcpp} package
\citep{Eddelbuettel.2011, Eddelbuettel.2013}.

\subsection{Source Points}
To initiate virtual dispersers at biologically meaningful source points, we
randomly placed source points within contiguous protected areas larger \(>\) 700
km\textsuperscript{2} (\Cref{SourcePoints} (a)). This conforms to the average
home range requirement of resident wild dogs \citep{Pomilia.2015} and allowed us
to remove areas too small to host viable wild dog populations. By distributing
source points randomly, the number of source points per km\textsuperscript{2}
within protected areas was approximately equal across our study area. Across all
protected areas, we distributed 50'000 source points, each representing the
starting point of a single dispersal trajectory. To render potential immigrants
into the study system, we placed additional 30'000 source points inside the
buffer zone around the main study area (\Cref{SourcePoints} (b)).

\begin{figure}[htbp]
  \begin{center}
    \includegraphics[width = \textwidth]{99_SourceAreas.png}
    \caption{(a) Illustration of an artificially expanded covariate layer (water
    cover). We expanded the layer by adding a 100 km buffer (light blue) which
    we filled with values sampled from the original layer. (b) Different source
    areas from which we released virtual dispersers. We initiated 50'000
    dispersers within the main study area (green) and another 30'000 dispersers
    within a virtual buffer (blue).}
    \label{SourcePoints}
  \end{center}
\end{figure}

\subsection{Heatmaps}
To identify dispersal hotspots across our study area, we rasterized all
simulated trajectories and created a heatmap indicating the absolute frequency
at which each raster-cell in the study area was visited by our virtual
dispersers. If the same trajectory crossed a raster-cell twice, it was only
counted once. That is, we did not consider revisits, thereby mitigating biases
arising from individuals that were trapped in a certain locations and moved
around in circles. We achieved high performance rasterization of all simulated
trajectories using the R-package {\tt terra} \citep{Hijmans.2020}. While one
could easily subset the data to visualize heatmaps assuming different dispersal
durations, we retained all 2'000 simulated steps to create a single heatmap.

% To examine if and how ``heat'' changes in response to changes in the location of
% source points and the number of simulated steps, we followed a 2 x 6 design and
% created heatmaps for both point sampling regines, as well as for 68, 125, 250,
% 500, 1000, and 2000 dispersal steps. We quantified the similarity of the
% resulting 12 heatmaps to the permeability and least-cost corridor maps presented
% in (Hofmann ...) we used Bhattacharyya's affinity. Bhattacharyya's affinity
% ranges from zero (complete separation) to one (perfect match) and has earlier
% been proposed to compare the overlap of utilisation distributions (Fieberg).

\subsection{Network Analysis I}
To pinpoint areas that are of high relevance for connecting remote regions in
our study area, we adopted a network view on the simulated trajecotries and
computed the betweenness metric. More specifically, we overlaid the study area
(including the buffer) with a with a raster that had a cell size of 5 x 5 km,
where each raster-cell represented a node in the final network. We then used the
simulated trajectories to determine all transitions occurring from one cell to
another, as well as the frequency at which these transitions occurred. This
resulted in an edge-list that we translated into a weighted network using the
r-package {\tt igraph}. Because {\tt igraph} handles the weights \(\omega\) as
costs, we inverted the traversal frequency at each cell by applying \(\omega =
\frac{\sum_i^n{Traversal Frequency_i}/n}{Traversal Frequency_i}\). Finally, we
calculated the betweenness metric for each node in the final network. This
metric indicates how often a specific raster-cell lies on a shortest path
between two other raster-cells and is therefore a useful metric to detect
movement corridors \citep{BastilleRousseau.2018}.

\subsection{Network Analysis II}
Besides this, we also adopted an alternative network view and identified the
frequency at which dispersers originating from one national park successfully
moved into an other national park. We achieved this by looking at each pairwise
combination of national parks and determining the relative number of
trajectories that originated in one national park and had at least one
subsequent coordinate laying in the other national park. This allowed us to
determine the average duration it took a simulated disperser to move from one
national park to another, and to point out combinations of national parks
between no links were observed.
